Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-17 draft.tex
Revised version

% \abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.}
\abstract{ 
The goal of epidemiology is to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess the predictive quality of image-derived variables for lower back pain.
%%
Therefore, an interactive visual analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as the spine canal course, its curvature and torsion.
%%
We analyze the predictive power of shape-describing variables for non-image variables, e.g., pain indicators.
%%
Therefore, we automatically extract a decision tree classifier for each non-image variable.
%%
Furthermore, we provide an IVA technique to compare the classifiers with interactive scatterplot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors.
%%
A correlation between lumbar spine shape and lower back pain disease was not existent in the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, correlations between non-image variables were successfully detected and characterized.
}

@@

\section{\uppercase{Introduction}}
\label{sec:Introduction}
%%Epidemiology is the study of causation of diseases.
Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP}, gather as much information as possible about participants to be assessed towards different diseases.
%%
These information are used to determine risk factors for diseases, help people to improve their lifestyle and their health or to support the diagnosis of widespread diseases.
%%
So far, epidemiological research is strongly hypothesis-driven.
%%
More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
%%
As a result, the hypothesis will be accepted or rejected.
%%

@@

In this paper, we present a thorough analysis of the image-derived data of a cohort study including lumbar back spine datasets.
%%
We analyze their discriminating power for different characteristics of back pain.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
The workflow is part of an interactive visual analysis (IVA) framework, constructed to enhance the epidemiological workflow with iterative data driven techniques.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis,
%%
\item The detection of correlations between image-based and socio-demographic as well as medical variables for an improved hypothesis generation,
%%
\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
%%
\end{itemize}
%%
The analysis, as well as the presented IVA tool is available as supplementary material in a interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
%%
%The IVA tool and all shown plots presented in this paper, except of Fig.~\ref{fig:age-gender}, are available as supplementary material under the following link: \url{blind.dnsalias.com}.
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To prove a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
For example, the analysis of the influence of nutrition on prostate cancer is strongly age dependent and requires therefore age-normalization.
%%
%When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
%\subsection{Epidemiological Data}
\subsection{Challenges of Epidemiological Data Analysis}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies analyzing one specific disease and its influence on the human body.
%%
We focus on data from large scale cohort studies.
%%
These studies aim at collecting as much data as possible for each subject.
%%
As a result, these data can be analyzed regarding many diseases and conditions.
%%

@@

Epidemiological data acquisition includes a variety of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
%Epidemiologists acquire data using a wide range of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
%%
This yields a heterogenous information space.
%%
To compare these data, information reduction techniques are necessary.
%%
For example, continuous data, such as age is often discretized into age-bins.
%%
Every information reduction can introduce a bias to the data, since it proposes an assumption about the data.
%%
Using age to divide subjects into \emph{young} and \emph{old} categories can distort statistical results.
%%
This distortion is reduced with increasing number of discretization steps.
%%

Modern cohort studies often comprise medical image data.
%%
These data are hard to analyze.
%%
Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be carefully adapted for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
%The lower (\emph{lumbar}) spine is the most stressed spine part .
%%
Lower back pain is one of the most frequent diseases in the western civilization \cite{Hoy2010}.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, sporting activities and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are not known.
%%
Potential \emph{confounding} effects are also subject of current research.
%%

@@

\cite{Klemm2014VIS} propose an extension to the previously described epidemiological workflow using \emph{Interactive Visual Analysis} (IVA).

@@

The effort of analyzing the decision trees increases with decision tree complexity.

@@

Gla{\ss}er et al. use variables derived from DCE-MRI data capturing the perfusion in the tumorous tissue \cite{Glasser2013}.

@@

Niemann et al. assess hepatic steatosis (fatty liver disease) risk factors using decision trees \cite{Niemann2014}.

@@

Pinhero et al. applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data \cite{Pinheiro}.
%%
Sekhavat and Hoeber propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map \cite{Sekhavat}.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial value and a target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
Klemm et al. analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects \cite{Klemm2013VMV}.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier clusters.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
Our data extraction process is based on the detection algorithm and lumbar spine canal extraction presented by \cite{Klemm2013VMV}.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures.
%%
%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task \cite{Klemm2014BVM}.
%%
\\\\
Unique for our work is the combination of data mining techniques with an IVA-approach by observing interesting feature relations in the context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
%%
\section{\uppercase{The Lumbar Spine Data Set}}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows the simultaneous analysis of many variables.
%%
Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}) including non-image and image-derived data.
%%
%It is an active research area to determine these correlations.
%%

%%
\subsection{Non-Image Data}
%\com{Analysis Using Heterogeneous Correlation}
The non-image variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors, e.g. sporting activity or nutrition.
%%
%Without the 9 image-derived parameter, the data set comprises of 134 variables:
The data set comprises of 134 variables (and 9 additional image-derived variables):
\begin{itemize}
	\item 21 metric variables, mostly describing somatometric variables and markers retrieved from blood analysis
	\item 113 categorical variables divided into
	\begin{itemize}
		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. pancreatitis or high blood pressure
		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
	\end{itemize}
\end{itemize}
%%
Many variables comprise missing data.
%%
Some variables are follow-up questions, covering the treatment of a disease.
%%
Others are exclusive to subgroups, such as women-specific variables, e.g. menstrual status.
%%
Sparse variables are statistically less resilient and need to be treated with care during the analysis.

@@

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model is initialized using three user-defined landmarks.
%%
A click on the L3 vertebra center initialized the model, two clicks on the top and bottom of the vertebra describes an initial height estimation (Fig.~\ref{fig:centerline}~a).
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae.
%%
Each subjects' model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, 
rotation and scaling.
%%
The detection fails for several subjects due to imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ out of $6,753$ lumbar spine models were obtained and constitute the foundation of this work.
%%
% \com{From VIS'14 Paper}
% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
% %%
% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
% %%
% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
% %%
% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
% %%
% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
% %%
% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
% %%
% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
% %%
% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
% %%

@@

\section{\uppercase{Experiments}}

@@

Larger people have also a longer spine and its shape is more straight.

@@

When correlating spine curvature and torsion with non-image variable, all these variables need to be taken into account

@@

Since the gender confounder mainly encodes body-size, we decided to divide subjects into body-size groups.

@@

\subsection{Preliminary Results}

@@

Pairwise combinations of image-derived variables are visualized via scatter plots left of the matrix diagonal and their correlation with back pain is denoted right of the matrix diagonal.

@@

\noindent As a first experiment we correlated the shape variable with the binary back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.

@@

\subsubsection{Correlation Matrix}

@@

Also the mean curvature and the curvature angle correlate by a factor of $-0.89$.

@@

Torsion does not correlate with any other image-derived variable.

@@

\subsubsection{Correlation of Image Variables With Back Pain}

@@

\subsubsection{Assessing the Information Gain Using the Principal Component Analysis}
%%
To determine the information gain per image-derived variables, we calculated a Principal Component Analysis (PCA) and compared the loadings per dimension.
%%
The PCA results in a linear projection of the variable space ordering their components by maximum variance.
%%
The first three principal components explain 75\% of the variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of the variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of the variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%%
The complete analysis can be found under the experiments section of the supplementary material (\url{blind.dnsalias.com})
%% ------------------------------------------
\subsubsection{Heterogenous Correlations}
%%
We then expanded our focus on correlations of image-derived variables with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
%%
The method uses the following correlation metrics for the different type combinations:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - perfect correlation}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
These variables have to be omitted in the analysis, since they are not statistically resilient.
%%
We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1, as presented in \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the variable (see \url{blind.dnsalias.com}).
%%
Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
%%

These observations brought us to the conclusion to incorporate more sophisticated data mining techniques to assess the influence of the image-derived variables.

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal curvature and torsion.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variable explain non-image variables.
%%
Decision trees are a popular classification method in data mining to create predictive models.
%%
Leafs of a decision tree represent class labels, branches represent decision trees.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
This allows epidemiologists to interpret the results without having deep knowledge about the algorithm creating the tree.
%%
Readability is only granted for small trees. %, complex structures with many branches are not desirable.
%%
Too many branches also impose overfitting to the data \cite{DecisionTree}.

\subsection{The C4.5/C5.0 Algorithm}
The C4.5 algorithm builds decision trees based on information entropy on a data set.
%%
Such a calculation requires a numeric or categorical target variable.
%%
%The algorithm then tries to find a decision tree, which divides the samples using the input variables just like the target variables.
%%
%This means, that every node in the tree represents the attribute which splits the data most efficiently into the target subset.
%%
The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
\begin{algorithm}[]
 Check for base cases\;
 \For{each attribute $a$}{
  Find the normalized information gain ratio from splitting on $a$\;
 }
 Let $a_{best}$ be the attribute with the highest normalized information gain\;
 Create a decision node that splits on $a_{best}$\;
 Recurse on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
 \caption{Building a decision tree using the C4.5 Algorithm}
 \label{alg:c4.5}
\end{algorithm}
C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%%
We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
%%
% \begin{enumerate}
% %%
% \item Check for base cases
% %%
% \item For each attribute $a$
% %%
% \begin{enumerate}
% %%
% \item Find the normalized information gain ratio from splitting on $a$
% %%
% \end{enumerate}
% %%
% \item Let $a_{best}$ be the attribute with the highest normalized information gain
% %%
% \item Create a decision node that splits on $a_{best}$
% %%
% \item Recurse on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node
% %%
% \end{enumerate}
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
Creating dummy variables bypasses this problem.
%%
%
%The actual use for the resulting tree is the classification of new observations (subjects).
%%
%Yet, we are interested in the decision rules and the classification accuracy.
%%
In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.

>>>>

% \abstract{The abstract should summarize the contents of the paper and should contain at least 70 and at most 200 words.}
\abstract{ 
The goal of epidemiology is to provide insight into disease causations.
%%
Hence, subject groups (cohorts) are analyzed to correlate the subjects' varying lifestyles, their medical properties and diseases.
%%
Recently, these cohort studies comprise medical image data.
%%
We assess the predictive quality of image-derived variables for lower back pain.
%%
Therefore, an Interactive Visual Analysis (IVA) framework was created and tested with 2,540 segmented lumbar spine data sets.
%%
The segmentation results are evaluated and quantified by employing shape-describing variables, such as the spine canal course, its curvature and torsion.
%%
We analyze the predictive power of shape-describing variables for non-image variables, e.g., pain indicators.
%%
Therefore, we automatically extract a decision tree classifier for each non-image variable.
%%
Furthermore, we provide an IVA technique to compare the classifiers with interactive scatterplot visualizations.
%%
As a first result, we conclude that image-based variables are only sufficient for classification of lifestyle factors.
%%
A correlation between lumbar spine shape and lower back pain disease was not existent in the automatically trained classifiers.
%%
However, the presented approach is a valuable extension for the IVA of epidemiological data.
%%
Hence, correlations between non-image variables were successfully detected and characterized.
}

@@

\section{\uppercase{Introduction}}
\label{sec:Introduction}
%%Epidemiology is the study of causation of diseases.
\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP}, gather as much information as possible about participants to be assessed towards different diseases.
%%
These information are used to determine risk factors for diseases, help people to live healthier or to support the diagnosis of widespread diseases.
%%
So far, epidemiological research is strongly hypothesis-driven.
%%
More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
%%
As a result, the hypothesis will be accepted or rejected.
%%

@@

We present a thorough analysis of the image-derived data of a cohort study including lumbar spine datasets.
%%
We analyze their discriminative power for different back pain characteristics.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
%Then, interactive visual analysis provides insight into the quality of the image-derived data as well as their potential for possible back pain risk factors.
Then, Interactive Visual Analysis (IVA) provides insight into the quality of the image-derived data as well as their potential for being back pain risk factors.
%%
The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data driven techniques.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
%%
\item The identification and evaluation of lumbar spine shape properties for back pain diagnosis,
%%
\item The detection of correlations between image-based and socio-demographic as well as medical variables for an improved hypothesis generation,
%%
\item The identification of the most important variables via data mining methods (including a novel semi-quantitative evaluation concept) embedded in our IVA-framework
%%
\end{itemize}
%%
The analysis, as well as the presented IVA tool is available as supplementary material in a interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
%%
%The IVA tool and all shown plots presented in this paper, except of Fig.~\ref{fig:age-gender}, are available as supplementary material under the following link: \url{blind.dnsalias.com}.
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item To prove a hypothesis, epidemiologists compile a list of variables depicting it.
%%
	\item Statistical methods, such as regression analysis, assess the association of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age.
%%
For example, the effect of nutrition on prostate cancer requires age-normalization due to the increased incidence for older subjects.
%%
%When for example analyzing influences of nutrition to prostate cancer, the results need to be age-normalized.
%%
Age acts as a \emph{confounder} for prostate cancer.
%%
%These \emph{confounding} variables are often hard to find.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
%\subsection{Epidemiological Data}
\subsection{Challenges of Epidemiological Data Analysis}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies analyzing one specific disease and its influence on the human body.
%%
We focus on large scale cohort study data.
%%
These studies aim at collecting as much data as possible for each subject.
%%
These data can be analyzed regarding many diseases and conditions.
%%

@@

Epidemiological data acquisition includes a variety of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
%Epidemiologists acquire data using a wide range of techniques, such as medical examinations, self reported questionnaires or genetic examinations.
%%
This yields a heterogenous information space.
%%
To compare these data, information reduction techniques are necessary.
%%
For example, continuous data, such as age, is often discretized into quantile bins.
%%
%Every information reduction can introduce a bias to the data, since it proposes an assumption about the data.
Every information reduction proposes an assumption about the data and can therefore introduce a bias.
%%
For example, using age to divide subjects into \emph{young} and \emph{old} categories can distort statistical results.
%%
This distortion is reduced with increasing number of discretization steps.
%%

Modern cohort studies often comprise medical image data.
%%
These data are hard to analyze.
%%
Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be carefully adapted for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
%The lower (\emph{lumbar}) spine is the most stressed spine part .
%%
Lower back pain is one of the most frequent diseases in the western civilization \cite{Hoy2010}.
%%
Epidemiologists assume associations between lumbar back pain and lifestyle factors.
%%
These include nutrition, physical activity and body posture combined with physical stress at work.
%%
The exact causes as well as particularly vulnerable risk groups are unknown.
%%
Potential \emph{confounding} effects are also subject of current research.
%%

@@

\cite{Klemm2014VIS} propose an extension to the previously described epidemiological workflow using IVA.

@@

The effort of analyzing decision trees increases with their complexity.

@@

\cite{Glasser2013} use variables derived from DCE-MRI data capturing the perfusion in the tumorous tissue.

@@

\cite{Niemann2014} assess hepatic steatosis (fatty liver disease) risk factors using decision trees.

@@

\cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data.
%%
\cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial and a target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
\cite{Klemm2013VMV} analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier clusters.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
Our data extraction process is based on the detection algorithm and lumbar spine canal extraction presented by \cite{Klemm2013VMV}.
%%
Clustering techniques applied to non-image data are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
%%
%Klemm et al. found \texttt{k-Prototypes} and \texttt{DBSCAN} appropriate for this task.
%%
\\\\
Unique for our work is the combination of data mining techniques with an IVA-approach by observing interesting feature relations in the context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
%%
\section{\uppercase{The Lumbar Spine Data Set}}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows the simultaneous analysis of many variables.
%%
Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}) including non-image and image-derived data.
%%
%It is an active research area to determine these correlations.
%%

%%
\subsection{Non-Image Data}
%\com{Analysis Using Heterogeneous Correlation}
The non-image variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors, e.g. sporting activity or nutrition.
%%
%Without the 9 image-derived parameter, the data set comprises of 134 variables:
The data set comprises of 134 variables (and 9 additional image-derived variables described in the next subsection):
\begin{itemize}
	\item 21 metric variables, mostly describing somatometric variables and markers retrieved from blood analysis
	\item 113 categorical variables divided into
	\begin{itemize}
		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. pancreatitis or high blood pressure
		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
	\end{itemize}
\end{itemize}
%%
Many variables comprise missing data.
%%
Some variables are follow-up questions, covering the treatment of a disease.
%%
Others are exclusive to subgroups, such as women-specific variables, e.g. menstrual status.
%%
Sparse variables are statistically less resilient and need to be treated with care during the analysis.

@@

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model is initialized using three user-defined landmarks.
%%
A click on the L3 vertebra center initializes the model, two clicks on the top and bottom of the vertebra describes an initial height estimation (Fig.~\ref{fig:centerline}~a).
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae.
%%
Each subjects' model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, 
rotation and scaling.
%%
The detection fails for several subjects due to imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ out of $6,753$ lumbar spine models were obtained and constitute the foundation of this work.
%%
% \com{From VIS'14 Paper}
% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
% %%
% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
% %%
% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
% %%
% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
% %%
% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
% %%
% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
% %%
% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
% %%
% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
% %%

@@

\section{\uppercase{Experiments and Preliminary Results}}

@@

Larger people have also a longer spine with a straighter shape.

@@

All these variables need to be taken into account, when spine curvature and torsion is correlated with non-image variables.

@@

Since the gender confounder mainly encodes body size, we decided to divide subjects into body size groups.

@@

@@

Pairwise combinations of image-derived variables are visualized via scatter plots on the left of the matrix diagonal.
Their correlation with back pain is denoted right of the matrix diagonal.

@@

\paragraph{GPLOM Analysis.}
As a first experiment we correlated the shape variable with the dichotomous back pain indicator using \texttt{GPLOM's} \cite{GPLOMS}.

@@

\paragraph{Correlation Matrix.}

@@

Also the \emph{mean curvature} and \emph{curvature angle} correlate by a factor of $-0.89$.

@@

\emph{Torsion} does not correlate with any other image-derived variable.

@@

\paragraph{Correlation of Image Variables With Lumbar Back Pain.}

@@

\paragraph{Assessing the Information Gain Using the Principal Component Analysis.}
%%
To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
%%
The PCA results in a linear projection of the variable space ordering their components by maximum variance.
%%
The first three principal components explain 75\% of the variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of the variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of the variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%%
The complete analysis can be found under the experiments section of the supplementary material (\url{blind.dnsalias.com})
%% ------------------------------------------
\paragraph{Heterogenous Correlations.}
%%
We then expanded our focus on correlations of image-derived variables with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
%%
The method uses the following correlation metrics for the different type combinations:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - perfect correlation}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
These variables have to be omitted in the analysis, since they are not statistically resilient.
%%
We display the resulting \emph{contingency matrix} using a heat map, encoding correlation values with color brightness with white for $0$ and dark blue for $1$, as presented in \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the variable (see \url{blind.dnsalias.com}).
%%
Only weak correlations could be found for \emph{mean curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation of \emph{torsion} with \emph{parkinson} (0.24).
%%
Other than that, \emph{torsion} correlated with almost no variables (p values between 0 and 0.05).
%%

These observations brought us to the decision to incorporate sophisticated data mining techniques to assess the influence of the image-derived variables.

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variable explain non-image variables.
%%
Decision trees are a popular classification method in data mining to create predictive models.
%%
%Leafs of a decision tree represent class labels, branches represent decision trees.
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
Epidemiologists can interpret the results without having deep knowledge about the classification algorithm.
%%
Readability is only granted for small trees. %, complex structures with many branches are not desirable.
%%
Too many branches also impose overfitting to the data \cite{DecisionTree}.
